---
title: "Group 9 Project"
format:
  pdf:
    documentclass: article
    geometry:
      - margin=1.0cm
    fontsize: 10pt
    linestretch: 1.05
    fig-pos: "htbp"
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| echo: false
#| warning: false
#| message: false
library(ggplot2)
library(tidyverse)
library(sjPlot)
library(performance)
library(kableExtra)
library(knitr)
library(stringr)
DAProject9 <- read.csv("DAProject9.csv")
```


# Introduction {#sec-intro}

## Background

The *2008 - 2012 Scottish Health Survey* provides key insight into the health trends of Scottish residents. In particular, the **weight classification group** of individuals has been closely monitored, alongside an array of lifestyle factors. In this report, we seek to explore the prevalence of **obesity** in the population over time, as well as its relationship with these lifestyle factors.

## Variables

`AgeGroup`: Age range of individual.

`Sex`: Male/Female.

`Employment`: Employment status of individual.

`Year`: Year of survey data collection.

`BMIgroup`: Weight classification group of individual.

## Questions of Interest

1.  Has the prevalence of obesity in Scotland changed over the given years of the Scottish Health Survey?

2.  Are there any differences in obesity by age, gender, socio-economic status or lifestyle factors?

@sec-eda consists of an exploratory analysis of the Survey data and explores the potential relationship between obesity and time, and whether there are any differences in obesity by other factors. @sec-fda contains the results from fitting a multiple regression model to the data, as well as the assessment of the model assumptions. Concluding remarks are given in @sec-conc.

# Exploratory Data Analysis {#sec-eda}

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-barplot
#| fig-cap: Barplot showing the proportion of each BMI Group, grouped by year
ggplot(DAProject9, aes(x = factor(Year), fill = BMIgroup)) +
  geom_bar(position = "fill") +
  labs(
    title = "Proportion of BMI Group by Year",
    x = "Year", 
    y = "Proportion")
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-linegraph
#| fig-cap: Linegraph showing the trend of BMI Group, year on year
#| error: true
summary_table <- DAProject9 %>%
  group_by(Year, BMIgroup) %>%
  summarise(Count = n(), .groups = "drop_last") %>%
  mutate(
    Total_In_Year = sum(Count),
    Proportion = Count / Total_In_Year)
  
ggplot(summary_table, aes(x = factor(Year), y = Proportion, color = BMIgroup, group = BMIgroup)) +
  geom_line(linewidth = 1.2) + 
  geom_point(size=3) +
  labs(
    title = "Trend of BMI Groups",
    x = "Year", 
    y = "Proportion of Population",
    color = "BMI Group"
  )
```

As displayed in @fig-barplot and @fig-linegraph, the proportion of each *BMIgroup* category does not appear to have changed much between 2008-2012, based off of the data collected. @fig-linegraph shows that the proportion of *Underweight* responses has remained constant, whereas *Overweight*, *Obese*, and *Normal* have displayed small changes, but there does not appear to be a trend year-on-year.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-employmentprop
#| tbl-cap: Percentage of each BMIgroup category in each category of Employment

emp_summary <- DAProject9 %>%
  group_by(Employment, BMIgroup) %>%
  summarise(Count = n(), .groups = "drop_last") %>%
  mutate(
    Proportion = Count / sum(Count),
    Percentage = round(Proportion * 100, 1)
  ) %>%
  ungroup()
emp_table_wide <- emp_summary %>%
  select(Employment, BMIgroup, Percentage) %>%
  pivot_wider(
    names_from = BMIgroup, 
    values_from = Percentage, 
    values_fill = 0
  ) %>%
  mutate(across(-Employment, ~paste0(., "%")))

kable(emp_table_wide)

```

We see from @tbl-employmentprop that based off of the category of the variable *Employment*, the percentage of people in full-time education and obese or overweight is significantly lower, whereas the percentage of those with a normal BMI is much higher.

The table also shows that those unable to work and those retired appear to have a higher percentage in the overweight category compared to other levels of *Employment*. These same groups also show a lower percentage of normal BMI.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-geomcol
#| fig-cap: Obesity prevalence by age group and sex
#| error: true
age_summary_prop <- DAProject9 %>%
  group_by(AgeGroup, Sex) %>% 
  summarise(
    Total_People_In_Group = n(),
    Obese_Count = sum(BMIgroup=="Obese"), 
    Percentage = Obese_Count / Total_People_In_Group, 
    .groups = 'drop'
  )


ggplot(data = age_summary_prop, aes(x = AgeGroup, y = Percentage,fill=Sex)) +
  geom_col(position = "dodge") + 
  scale_y_continuous() + 
  labs(
    title = "Obesity prevalence by age group and sex",
    x = "Age group",
    y = "Obesity prevalence"
  ) +
  scale_y_continuous(labels = scales::percent)
```

@fig-geomcol suggests a clear age gradient in obesity prevalence. The proportion classified as obese increases from the youngest age groups into middle and early older adulthood, before falling slightly in the 75+ group. The differences between male and female sexes appear smaller than the differences shown across age groups, suggesting age may be a more important source of variation than sex.

# Formal Data Analysis {#sec-fda}

Formal analysis was carried out using multivariable logistic regression, with obesity status as the binary response. Following the project brief, a new response variable was created from `BMIgroup`: respondents in the Obese category were coded as obese, while all remaining respondents were coded as not obese. Year, age group, sex and employment status were all entered as categorical explanatory variables. Because Project 9 does not contain direct lifestyle measures, the formal model focuses on the available survey year, demographic and socio-economic variables, with employment status acting as the socio-economic indicator.

No interaction terms were included in the main model, because the primary aim was to estimate the adjusted main effects of survey year and subgroup characteristics in a parsimonious model. The fitted model can be written as

$$
\text{logit}\{P(Y_i = 1)\}
= \beta_0 + \text{Year}_i + \text{AgeGroup}_i + \text{Sex}_i + \text{Employment}_i,
$$

where $Y_i = 1$ indicates that respondent $i$ was classified as obese. The reference categories were set to 2008 for year, 16--24 for age group, Female for sex, and In paid employment, self-employed or on gov't training for employment status.

```{r}
library(tidyverse)
library(sjPlot)
library(knitr)
library(stringr)

dat <- read.csv("DAProject9.csv", stringsAsFactors = FALSE) |>
  mutate(
    obesity = if_else(BMIgroup == "Obese", 1L, 0L),
    Year = factor(Year, levels = c(2008, 2009, 2010, 2011, 2012)),
    AgeGroup = factor(
      AgeGroup,
      levels = c("16-24", "25-34", "35-44", "45-54", "55-64", "65-74", "75+")
    ),
    Sex = factor(Sex, levels = c("Female", "Male")),
    Employment = factor(
      Employment,
      levels = c(
        "In paid employment, self-employed or on gov't training",
        "Doing something else",
        "In full-time education",
        "Looking after home/family",
        "Looking for/intending to look for paid work",
        "Perm unable to work",
        "Retired"
      )
    )
  )

fit_obesity <- glm(
  obesity ~ Year + AgeGroup + Sex + Employment,
  data = dat,
  family = binomial
)
```

To assess the overall contribution of each explanatory variable, likelihood ratio tests were carried out by removing each term from the full model in turn. The results are shown below.

```{r}
#| tbl-cap: "Likelihood ratio tests for the multivariable logistic regression model"
#| label: tbl-likelihood

lrt_results <- drop1(fit_obesity, test = "Chisq") |>
  as.data.frame() |>
  tibble::rownames_to_column("Term") |>
  filter(Term != "<none>") |>
  transmute(
    Term,
    `LR chi-square` = round(LRT, 2),
    df = Df,
    `p-value` = ifelse(`Pr(>Chi)` < 0.001, "<0.001", sprintf("%.3f", `Pr(>Chi)`))
  )

kable(lrt_results, align = c("l", "r", "r", "r"))
```

As shown in @tbl-likelihood, Year did not make a statistically significant contribution to the adjusted model ($\chi^2 = 4.03$, $df = 4$, $p = 0.401$). Therefore, there is no strong adjusted evidence that obesity prevalence changed across the five survey years. In contrast, AgeGroup made by far the strongest contribution ($\chi^2 = 287.69$, $df = 6$, $p < 0.001$), while Employment also remained significant after adjustment ($\chi^2 = 66.06$, $df = 6$, $p < 0.001$). Sex was not significant overall ($\chi^2 = 1.65$, $df = 1$, $p = 0.199$). Taken together, these results suggest that the main differences in obesity prevalence were associated with age and employment status rather than survey year or sex.

The adjusted odds ratios provide the main interpretation of the fitted model. Exponentiated coefficients are presented with 95% confidence intervals to summarise the adjusted associations with obesity.

```{r}
#| tbl-cap: "Adjusted odds ratios from the multivariable logistic regression model"
#| label: tbl-adjusted

coef_mat <- summary(fit_obesity)$coefficients

or_results <- tibble(
  term = rownames(coef_mat),
  estimate = coef_mat[, 1],
  se = coef_mat[, 2],
  p_value = coef_mat[, 4]
) |>
  filter(term != "(Intercept)") |>
  mutate(
    OR = exp(estimate),
    CI_low = exp(estimate - 1.96 * se),
    CI_high = exp(estimate + 1.96 * se),

    Variable = case_when(
      str_starts(term, "Year") ~ "Year",
      str_starts(term, "AgeGroup") ~ "Age group",
      term == "SexMale" ~ "Sex",
      str_starts(term, "Employment") ~ "Employment",
      TRUE ~ term
    ),

    Level = case_when(
      str_starts(term, "Year") ~ str_remove(term, "^Year"),
      str_starts(term, "AgeGroup") ~ str_remove(term, "^AgeGroup"),
      term == "SexMale" ~ "Male",
      str_starts(term, "Employment") ~ str_remove(term, "^Employment"),
      TRUE ~ term
    ),

    Level = recode(
      Level,
      "Doing something else" = "Doing something else",
      "In full-time education" = "Full-time education",
      "Looking after home/family" = "Home/family",
      "Looking for/intending to look for paid work" = "Seeking paid work",
      "Perm unable to work" = "Unable to work",
      "Retired" = "Retired"
    ),

    `Odds ratio` = sprintf("%.2f", OR),
    `95% CI` = sprintf("%.2f to %.2f", CI_low, CI_high),
    `p-value` = ifelse(p_value < 0.001, "<0.001", sprintf("%.3f", p_value))
  ) |>
  select(Variable, Level, `Odds ratio`, `95% CI`, `p-value`)

knitr::kable(
  or_results,
  align = c("l", "l", "r", "r", "r"),
  booktabs = TRUE
)
```

As shown in @tbl-adjusted, the adjusted odds ratios showed a clear age gradient in obesity. Compared with respondents aged 16–24, those aged 55–64 and 65–74 had the highest odds of obesity, at 2.55 and 2.49 times respectively. All other age groups above 24 also had significantly higher odds. By contrast, there was no evidence of a sex difference after adjustment, with the odds ratio for males compared with females close to 1.

Employment status also had a significant effect on obesity. Compared with the reference group, people in full-time education had lower odds of obesity, while those permanently unable to work had higher odds. Other employment groups were not statistically significant because their odds ratios were close to 1. Overall, after adjusting for other variables, obesity was mainly associated with age group and employment status rather than sex. The year variable also showed no clear change over time.

```{r}
#| fig-cap: "Adjusted odds ratios for obesity from the multivariable logistic regression model."
#| fig-width: 8
#| fig-height: 3.8
#| out-width: "82%"
#| fig-align: "center"
#| label: fig-or
#| fig-pos: "H"
plot_model(
  fit_obesity,
  type = "est",
  transform = "exp",
  show.values = TRUE,
  show.p = FALSE,
  value.offset = 0.28,
  vline.color = "grey40",
  line.size = 0.9,
  dot.size = 2.5,
  title = "Odds ratios for obesity",
  axis.title = c("Odds ratio", ""),
  term.labels = c(
    "Year [2009]",
    "Year [2010]",
    "Year [2011]",
    "Year [2012]",
    "AgeGroup [25–34]",
    "AgeGroup [35–44]",
    "AgeGroup [45–54]",
    "AgeGroup [55–64]",
    "AgeGroup [65–74]",
    "AgeGroup [75+]",
    "Sex [Male]",
    "Employment [Doing something else]",
    "Employment [In full-time education]",
    "Employment [Looking after home/family]",
    "Employment [Looking for/intending to look for paid work]",
    "Employment [Perm unable to work]",
    "Employment [Retired]"
  )
) +
  theme_classic(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.y = element_text(size = 10)
  )
```

@fig-or shows the adjusted odds ratios for obesity from the multivariable logistic regression model, with 95% confidence intervals. After adjustment for the other variables, survey year showed little evidence of change in obesity prevalence, as all year-specific odds ratios were close to 1. In contrast, age group showed the strongest pattern, with respondents aged 55–64 and 65–74 having substantially higher odds of obesity than those aged 16–24. Employment status also remained important, with lower odds among those in full-time education and higher odds among those permanently unable to work. Sex was not associated with obesity after adjustment.

```{r}
#| fig-cap: "Adjusted predicted probability of obesity by survey year"
#| fig-width: 5.5
#| fig-height: 3.6
#| out-width: '65%'
#| fig-align: 'center'
#| label: fig-survey
#| fig-pos: "H"
plot_model(
  fit_obesity,
  type = "pred",
  terms = "Year",
  axis.title = c("Survey year", "Predicted probability of obesity")
) +
  theme(
    plot.title = element_blank()
  )
```

As shown in @fig-survey, the adjusted predicted probabilities of obesity were very similar across the five survey years, and the confidence intervals overlapped substantially. This visual pattern agrees with the likelihood ratio test and suggests that obesity prevalence remained broadly stable from 2008 to 2012 after controlling for age group, sex and employment.

```{r}
#| fig-cap: "Adjusted predicted probability of obesity by age group"
#| fig-width: 5.5
#| fig-height: 3.6
#| out-width: '65%'
#| fig-align: 'center'
#| label: fig-age
#| fig-pos: "H"
plot_model(
  fit_obesity,
  type = "pred",
  terms = "AgeGroup",
  axis.title = c("Age group", "Predicted probability of obesity")
) +
  theme(
    plot.title = element_blank()
    )
```

As shown in @fig-age Adjusted predicted probabilities were also plotted for age group. The fitted probabilities increase clearly from 16--24 into middle adulthood, reach their highest level around 55--64 and 65--74, and then fall slightly in the 75+ group. This confirms a strong age gradient in obesity prevalence and is consistent with the large adjusted odds ratios for the middle-aged and early older groups.

```{r}
#| fig-cap: "Adjusted predicted probability of obesity by employment status"
#| fig-width: 8
#| fig-height: 4.2
#| out-width: '68%'
#| fig-align: 'center'
#| label: fig-employment
#| fig-pos: "H"
p_emp <- plot_model(
  fit_obesity,
  type = "pred",
  terms = "Employment",
  axis.title = c("Employment status", "Predicted probability of obesity")
)

p_emp +
  scale_x_continuous(
    breaks = sort(unique(p_emp$data$x)),
    labels = c(
      "Paid employment /\nself-employed /\ngov't training",
      "Doing\nsomething else",
      "In full-time\neducation",
      "Looking after\nhome/family",
      "Seeking\npaid work",
      "Unable\nto work",
      "Retired"
    )
  ) +
  theme(
    plot.title = element_blank(),
    axis.text.x = element_text(size = 9, lineheight = 0.9)
  )
```

As shown in @fig-employment, a final adjusted probability plot was used to compare employment groups more clearly. The highest fitted probability occurs among respondents who were permanently unable to work, while the lowest fitted probability occurs among those in full-time education. Most of the remaining employment groups are much closer to the reference category after adjustment, which supports the conclusion that employment differences are present but are more selective than the age differences.

Overall, the formal analysis provides a clear answer to the two research questions. There is no strong adjusted evidence that obesity prevalence changed over time between 2008 and 2012. However, there are clear subgroup differences, especially by age group and employment status. Obesity was most strongly associated with middle and early older adulthood, and it was also more common among respondents permanently unable to work and less common among those in full-time education. After adjustment for the other variables, there was no clear overall sex difference. Because this dataset does not include direct lifestyle variables, the formal analysis cannot assess lifestyle effects directly and is therefore limited to the available demographic and socio-economic measures.


# Conclusions {#sec-conc}

Analysis showed no relationship between the probability of obesity and time, suggesting that there has been *no change* in the **prevalence** of obesity in Scotland over the period of the study. However, we did find differences in obesity by some of our other variables. Namely, **age** and **employment status** were shown to relate to probability of obesity.

Our analysis suggests that one's probability of obesity *increases* into middle adulthood and peaks at the **55-64** and **65-74** groups before *declining* for those in the **75+** group. Those in **full-time education** had significantly *lower* odds than the baseline category, while those who are **permanently unable to work** had *higher* odds. All other employment status categories were *not* found to be significant. The individuals' **sex** was found to have *no effect* with probability of obesity.

This study provides great framework for understanding obesity in our population and can be used as a platform for future research. Recording additional lifestyle factors in future surveys, as well monitoring obesity trends over a larger period of time, will help to further our understanding of the issue.
