---
title: "Group 9 Project"
number-sections: true
format: 
  html:
    embed-resources: true
    code-tools: true
editor_options: 
  chunk_output_type: console
theme: flatly
---

```{r}
#| echo: false
#| warning: false
#| message: false
library(ggplot2)
library(tidyverse)
library(sjPlot)
library(performance)
library(kableExtra)
#this is the name given to the .csv file. If anyone is making plots, use data=DAProject9
DAProject9 <- read.csv("DAProject9.csv")
```


# Introduction {#sec-intro}

The 2008 - 2012 Scottish Health Survey provides key insight into the health trends of Scottish residents. It provides information on *age*, *sex*, *employment status*, and *BMI* of people, sorted by the year the survey data was collected. The following report seeks to explore the relationship of obesity over time, exploring whether or not the prevalence of obesity is changing. The report also seeks to explore whether or not there any differences in obesity by age, gender, socio-economic status or lifestyle factors.

@sec-eda consists of an exploratory analysis of the Survey data and explores the potential relationship between obesity and time, and whether there are any differences in obesity by other factors. @sec-fda contains the results from fitting a multiple regression model to the data, as well as the assessment of the model assumptions. Concluding remarks are given in @sec-conc.

# Exploratory Data Analysis {#sec-eda}


```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-barplot
#| fig-cap: Barplot showing the proportion of each BMI Group, grouped by year
ggplot(DAProject9, aes(x = factor(Year), fill = BMIgroup)) +
  geom_bar(position = "fill") +
  labs(
    title = "Proportion of BMI Group by Year",
    x = "Year", 
    y = "Proportion")
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-linegraph
#| fig-cap: Linegraph showing the trend of BMI Group, year on year
#| error: true
summary_table <- DAProject9 %>%
  group_by(Year, BMIgroup) %>%
  summarise(Count = n(), .groups = "drop_last") %>%
  mutate(
    Total_In_Year = sum(Count),
    Proportion = Count / Total_In_Year)
  
ggplot(summary_table, aes(x = factor(Year), y = Proportion, color = BMIgroup, group = BMIgroup)) +
  geom_line(linewidth = 1.2) + 
  geom_point(size=3) +
  labs(
    title = "Trend of BMI Groups",
    x = "Year", 
    y = "Proportion of Population",
    color = "BMI Group"
  )
```
As displayed in @fig-barplot and @fig-linegraph, the proportion of each *BMIgroup* category does not appear to have changed much between 2008-2012, based off of the data collected. @fig-linegraph shows that the proportion of *Underweight* responses has remained constant, whereas *Overweight*, *Obese*, and *Normal* have displayed small changes, but there does not appear to be a trend year-on-year.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-employmentprop
#| tbl-cap: Percentage of each BMIgroup category in each category of Employment

emp_summary <- DAProject9 %>%
  group_by(Employment, BMIgroup) %>%
  summarise(Count = n(), .groups = "drop_last") %>%
  mutate(
    Proportion = Count / sum(Count),
    Percentage = round(Proportion * 100, 1)
  ) %>%
  ungroup()
emp_table_wide <- emp_summary %>%
  select(Employment, BMIgroup, Percentage) %>%
  pivot_wider(
    names_from = BMIgroup, 
    values_from = Percentage, 
    values_fill = 0
  ) %>%
  mutate(across(-Employment, ~paste0(., "%")))

kable(emp_table_wide, caption = "Percentage of BMI Group within each Employment Category")

```

We see from @tbl-employmentprop that based off of the category of the variable *Employment*, the percentage of people in full-time education and obese or overweight is significantly lower, whereas the percentage of those with a normal BMI is much higher.

The table also shows that those unable to work and those retired appear to have a higher percentage in the overweight category compared to other levels of *Employment*. These same groups also show a lower percentage of normal BMI.

# Formal Data Analysis {#sec-fda}

# Conclusions {#sec-conc}
